Comparing the Effectiveness of the Choropleth Map with a Hexagon Tile Map for Communicating Cancer Statistics

### 2020-01-12 paper/paper.Rmd
Added lineup fig

<!-- This alternative display is suggested for communicating distributions that relate to the population values or density of the population in each geographic unit. This study uses the lineup protocol for visual inference testing of the recognition of spatial distributions. The Australian Cancer Atlas explores spatial distributions of cancer incidence and the burden on Australian communities. The spatial distributions of the cancers explored in this atlas inspired the trend models presented to participants in this study.

Motivation
=============
!-- Find a new display for the Cancer Atlas --

## The Australian Cancer Atlas
-->

@@

The paper is organised as follows. The next section discusses the background of geographic data display and visual inference procedures. The [Methodology] Section describes the methods for conducting the experiment and analysing the results. The results are summarized in Section [Results].

<!--
The Atlas examines Australian communities at the Statistical Areas at Level 2 (SA2) [@abs2016] used by the Australian Bureau of Statistics. Bayesian spatial smoothing was applied to integrate the values of neighboring areas. Spatial smoothing protects the privacy and gives stability to the estimated amounts in each SA2. Users of the Atlas can explore the difference of the statistic of each SA2 from the Australian average, for both diagnoses (Standardised Incidence Rates) and excess deaths. It communicates the value of the statistic for each SA2 geographic unit through color, by filling the areas using the diverging color scheme ranging from dark blue to red. 
-->

Background
=============

@@

<!-- Hexagon tile map -->
The hexagon tile map algorithm, autmatically matches spatial regions to their nearest hexagon tile, from a grid of tiles. It has the effect of spreading out the inner city areas while maintaining the spatial locations or regions in remote areas. The algorithm is available in the R package, sugarbag (XXX REF). Figure \ref{fig:liver} shows the hexagon tile map, along with the choropleth map of liver cancer rates in Australia. Colour maps from substantially below average (blue) to substantially above average (red) rates. The inner city areas are expanded out, making it possible to see the cancer incidence in the small, densely populated areas. Remote regions are represented by isolated hexagons, which is not ideal, but maintains the spatial location of these data values. It is oof interest to know how well the spatial distribution is perceived for this display, in comparison to the choropleth.

@@

The chance that an observer picks the data plot out of a lineup of size $m$ plots accidentally, if the null hypothesis is true is $/m$. With $K$ observers, the probability of $k$ randomly choosing the data plot, roughly follows a binomial distribution with $p=1/m$. Figure XXX (this needs to be placed near here) shows a lineup of the hexagon tile map, of size $m=12$. Plot XXX is the data plot, and the remaining 11 are plots of null data.

In order to determine the effectiveness of a type of display, this probability is less relevant than the overall proportion of observers who pick the data plot, $k/K$. The power of the test statistic (data plot) is provided by this proportion. Power in a statistical sense is the ability oof the statistic to *produce a rejection* of the null hypothesis, if it is indeed *not true*. With the same data plotted using two different displays, the display with the highest proportion of people who choose the data plot would be considered to be the most powerful statistic.

@@

The primary factor in the experiment is the plot type. The secondary factor is a trend model. Three trend models were developed, one mirroring a large spatial trend for which the choropleth would be expected to do well, and two with differing level of inner city hot spots. These latter two reflect the structure seen in the liver cancer data (Figure \ref{fig:liver}). This produces six treatment levels:

@@

Data is generated for each of the trend models, with four replicates, and each displayed both as a choropleth and as a hexagon tile map, which yields 12 data sets, and 24 data plots. This set of displays is divided in half, providing two sets of 12 displays, Group A and Group B. Participants were randomly allocated to Group A or B. Participants saw a data set only once, with as a choropleth or as a hexagon tile map. Table \ref{tab:exp-design} summarise the design. (XXX Maybe replace this table with your diagram illustrating the design??)

<!-- Each group did not see the same data for both map types. We generated four simulated sets of data for each treatment.
This will generate twenty-four lineups (twelve lineups of geographic maps and twelve lineups of hexagon tile maps). Participants evaluated twelve lineups, size of each map type.
For each of the six geographic displays and six hexagon displays, two of each trend model were shown to the participants. We outline the design in Table \ref{tab:exp-design}.
-->

```{r exp-design, results = "asis"}
d %>% 
  mutate(Trend = trend,
    Group = group,
    `Map Type` = type) %>% 
  group_by(Group, Trend, `Map Type`) %>% 
  summarise(Reps = paste0(sort(unique(replicate)), collapse = ", ")) %>% 
  pivot_wider(names_from = c("Map Type"), values_from = c("Reps")) %>% 
  ungroup() %>% 
  knitr::kable(., format = "latex", booktabs = TRUE, 
    linesep = c("", "", "\\addlinespace"),
    caption = "The experimental design results in the following allocation of replicate data sets within each trend model to groups A and B.") %>% kable_styling(bootstrap_options = "hold_position") %>% 
  collapse_rows(., columns = 1, custom_latex_hline = c(1:3))
```

@@

Fig. \ref{fig:detect-compare} displays the average detection rates for the two types of plot separately for each trend model. Each trend model was tested using four repetitions, evaluations on the same data set were seen as either choropleths or hexagon tile maps by each group as specified in Table. \ref{tab:exp-design}; the detection rates for each display are connected by a line segment. The Three Cities and All Cities trend models shown in the hexagon tile map allowed viewers to detect the data plot substantially more often than the choropleth counterparts. One replicate for the All Cities group, had similar detection rates for both plot type. Surprisingly, participants could also detect the gradual spatial trend in the NW-SE group from the hexagon tile map. We expected that the choropleth map would be superior for the type of spatial pattern, but the data suggests the hexagon tile map performs slightly better, or equally as well.

>>>>

@@

The paper is organised as follows. The next section discusses the background of geographic data display and visual inference procedures. The [Methodology] Section describes the methods for conducting the experiment and analysing the results. The results are summarized in Section [Results]. 
Background
=============

@@

<!-- Hexagon tile map -->
The hexagon tile map algorithm, autmatically matches spatial regions to their nearest hexagon tile, from a grid of tiles. It has the effect of spreading out the inner city areas while maintaining the spatial locations or regions in remote areas. The algorithm is available in the R package, sugarbag [@sugarbag]. Figure \ref{fig:liver} shows the hexagon tile map, along with the choropleth map of liver cancer rates in Australia. Colour maps from substantially below average (blue) to substantially above average (red) rates. The inner city areas are expanded out, making it possible to see the cancer incidence in the small, densely populated areas. Remote regions are represented by isolated hexagons, which is not ideal, but maintains the spatial location of these data values. It is oof interest to know how well the spatial distribution is perceived for this display, in comparison to the choropleth.

@@

The chance that an observer picks the data plot out of a lineup of size $m$ plots accidentally, if the null hypothesis is true is $1/m$. With $K$ observers, the probability of $k$ randomly choosing the data plot, roughly follows a binomial distribution with $p=1/m$. Fig. \ref{fig:lineup} shows a lineup of the hexagon tile map, of size $m=12$. Plot 3 is the data plot, and the remaining 11 are plots of null data.

```{r lineup, fig.cap = "This lineup of twelve hexagon tile map displays contains one map with a real population related structure. The rest are null plots that contain spatial correlation between neighbours.", fig.height = 8, fig.width = 8}
ggdraw() +
  draw_plot(rasterGrob(readPNG("figures/aus_cities_3_geo.png")))
```

In order to determine the effectiveness of a type of display, this probability is less relevant than the overall proportion of observers who pick the data plot, $k/K$. The power of the test statistic (data plot) is provided by this proportion. Power in a statistical sense is the ability of the statistic to *produce a rejection* of the null hypothesis, if it is indeed *not true*. With the same data plotted using two different displays, the display with the highest proportion of people who choose the data plot would be considered to be the most powerful statistic.

@@

The primary factor in the experiment is the plot type. The secondary factor is a trend model. Three trend models were developed, one mirroring a large spatial trend for which the choropleth would be expected to do well, and two with differing level of inner city hot spots. These latter two reflect the structure seen in the liver cancer data (Fig. \ref{fig:liver}). This produces six treatment levels:

@@

Data is generated for each of the trend models, with four replicates, and each displayed both as a choropleth and as a hexagon tile map, which yields 12 data sets, and 24 data plots. This set of displays is divided in half, providing two sets of 12 displays, Group A and Group B. Participants were randomly allocated to Group A or B. Participants saw a data set only once, with as a choropleth or as a hexagon tile map. Table \ref{fig:exp-design} summarises the design. (XXX Maybe replace this table with your diagram illustrating the design??)

```{r exp-design, results = "asis", fig.width = 8, fig.height = 8}
ggdraw(xlim = c(0,1), ylim = c(0,1)) + draw_plot(rasterGrob(png::readPNG("figures/experiment_design.png")))
```

@@

Fig. \ref{fig:detect-compare} displays the average detection rates for the two types of plot separately for each trend model. Each trend model was tested using four repetitions, evaluations on the same data set were seen as either choropleths or hexagon tile maps by each group as specified in Table. \ref{fig:exp-design}; the detection rates for each display are connected by a line segment. The Three Cities and All Cities trend models shown in the hexagon tile map allowed viewers to detect the data plot substantially more often than the choropleth counterparts. One replicate for the All Cities group, had similar detection rates for both plot type. Surprisingly, participants could also detect the gradual spatial trend in the NW-SE group from the hexagon tile map. We expected that the choropleth map would be superior for the type of spatial pattern, but the data suggests the hexagon tile map performs slightly better, or equally as well.

### 2020-01-12 paper/paper.tex
Added lineup fig

The paper is organised as follows. The next section discusses the background of geographic data display and visual inference procedures. The \protect\hyperlink{methodology}{Methodology} Section describes the methods for conducting the experiment and analysing the results. The results are summarized in Section \protect\hyperlink{results}{Results}.

\hypertarget{background}{%
\section{Background}\label{background}}

@@

The hexagon tile map algorithm, autmatically matches spatial regions to their nearest hexagon tile, from a grid of tiles. It has the effect of spreading out the inner city areas while maintaining the spatial locations or regions in remote areas. The algorithm is available in the R package, sugarbag (XXX REF). Figure \ref{fig:liver} shows the hexagon tile map, along with the choropleth map of liver cancer rates in Australia. Colour maps from substantially below average (blue) to substantially above average (red) rates. The inner city areas are expanded out, making it possible to see the cancer incidence in the small, densely populated areas. Remote regions are represented by isolated hexagons, which is not ideal, but maintains the spatial location of these data values. It is oof interest to know how well the spatial distribution is perceived for this display, in comparison to the choropleth.

@@

The chance that an observer picks the data plot out of a lineup of size \(m\) plots accidentally, if the null hypothesis is true is \(/m\). With \(K\) observers, the probability of \(k\) randomly choosing the data plot, roughly follows a binomial distribution with \(p=1/m\). Figure XXX (this needs to be placed near here) shows a lineup of the hexagon tile map, of size \(m=12\). Plot XXX is the data plot, and the remaining 11 are plots of null data.

In order to determine the effectiveness of a type of display, this probability is less relevant than the overall proportion of observers who pick the data plot, \(k/K\). The power of the test statistic (data plot) is provided by this proportion. Power in a statistical sense is the ability oof the statistic to \emph{produce a rejection} of the null hypothesis, if it is indeed \emph{not true}. With the same data plotted using two different displays, the display with the highest proportion of people who choose the data plot would be considered to be the most powerful statistic.

@@

Data is generated for each of the trend models, with four replicates, and each displayed both as a choropleth and as a hexagon tile map, which yields 12 data sets, and 24 data plots. This set of displays is divided in half, providing two sets of 12 displays, Group A and Group B. Participants were randomly allocated to Group A or B. Participants saw a data set only once, with as a choropleth or as a hexagon tile map. Table \ref{tab:exp-design} summarise the design. (XXX Maybe replace this table with your diagram illustrating the design??)

\caption{\label{tab:exp-design}The experimental design results in the following allocation of replicate data sets within each trend model to groups A and B.}
\centering
\begin{tabular}[t]{llll}
\toprule
Group & Trend & Choro. & Hex.\\
\midrule
 & NW-SE & 1, 2 & 3, 4\\
\cmidrule{2-4}
 & Three Cities & 1, 2 & 3, 4\\
\cmidrule{2-4}
\multirow{-3}{*}{\raggedright\arraybackslash A} & All Cities & 2, 4 & 1, 3\\
\cmidrule{1-4}
 & NW-SE & 3, 4 & 1, 2\\
\cmidrule{2-4}
 & Three Cities & 3, 4 & 1, 2\\
\cmidrule{2-4}
\multirow{-3}{*}{\raggedright\arraybackslash B} & All Cities & 1, 3 & 2, 4\\
\bottomrule
\end{tabular}
\end{table}

>>>>

The paper is organised as follows. The next section discusses the background of geographic data display and visual inference procedures. The \protect\hyperlink{methodology}{Methodology} Section describes the methods for conducting the experiment and analysing the results. The results are summarized in Section \protect\hyperlink{results}{Results}.
Background
=============

@@

The hexagon tile map algorithm, autmatically matches spatial regions to their nearest hexagon tile, from a grid of tiles. It has the effect of spreading out the inner city areas while maintaining the spatial locations or regions in remote areas. The algorithm is available in the R package, sugarbag {[}\protect\hyperlink{ref-sugarbag}{1}{]}. Figure \ref{fig:liver} shows the hexagon tile map, along with the choropleth map of liver cancer rates in Australia. Colour maps from substantially below average (blue) to substantially above average (red) rates. The inner city areas are expanded out, making it possible to see the cancer incidence in the small, densely populated areas. Remote regions are represented by isolated hexagons, which is not ideal, but maintains the spatial location of these data values. It is oof interest to know how well the spatial distribution is perceived for this display, in comparison to the choropleth.

@@

The chance that an observer picks the data plot out of a lineup of size \(m\) plots accidentally, if the null hypothesis is true is \(1/m\). With \(K\) observers, the probability of \(k\) randomly choosing the data plot, roughly follows a binomial distribution with \(p=1/m\). Fig. \ref{fig:lineup} shows a lineup of the hexagon tile map, of size \(m=12\). Plot 3 is the data plot, and the remaining 11 are plots of null data.

\begin{figure}
\centering
\includegraphics{paper_files/figure-latex/lineup-1.pdf}
\caption{\label{fig:lineup}This lineup of twelve hexagon tile map displays contains one map with a real population related structure. The rest are null plots that contain spatial correlation between neighbours.}
\end{figure}

In order to determine the effectiveness of a type of display, this probability is less relevant than the overall proportion of observers who pick the data plot, \(k/K\). The power of the test statistic (data plot) is provided by this proportion. Power in a statistical sense is the ability of the statistic to \emph{produce a rejection} of the null hypothesis, if it is indeed \emph{not true}. With the same data plotted using two different displays, the display with the highest proportion of people who choose the data plot would be considered to be the most powerful statistic.

@@

Data is generated for each of the trend models, with four replicates, and each displayed both as a choropleth and as a hexagon tile map, which yields 12 data sets, and 24 data plots. This set of displays is divided in half, providing two sets of 12 displays, Group A and Group B. Participants were randomly allocated to Group A or B. Participants saw a data set only once, with as a choropleth or as a hexagon tile map. Table \ref{fig:exp-design} summarises the design. (XXX Maybe replace this table with your diagram illustrating the design??)

\includegraphics{paper_files/figure-latex/exp-design-1.pdf}
